Live Programming with Typed Holes

### 2017-09-26 calculus.tex
fixed definition of substitution
d err judgement and fixes to other auxiliary judgments in the dynamics
fixes to auxiliary judgements and added proper rules for marks
step rule names
context membership tweaks per discussion with ian last week
Adding missing rule for substituion into casts
texed up rules for contextual dynamics (minus rule names)
added missing mark transitions and related rules
updating with most of the rule names, pending a chat with @cyrus- for a couple more
adding some possible theorem statements, ideas from Slack

\inferrule[ESVar]{ }{
  \expandSyn{\hGamma, x : \htau}{x}{\htau}{x}{\emptyset}
}

@@

\inferrule[TAVar]{ }{
	\hasType{\Delta}{\hGamma, x : \htau}{x}{\htau}
}

@@

\inferrule[TAEHole]{
  \hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}
}{
  \hasType{\Delta, \Dbinding{u}{\hGamma'}{\htau}}{\hGamma}{\dehole{u}{\sigma}{\markname}}{\htau}
}

\inferrule[TANEHole]{
  \hasType{\Delta}{\hGamma}{\dexp}{\htau'}\\
  \hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}
}{
  \hasType{\Delta, \Dbinding{u}{\hGamma'}{\htau}}{\hGamma}{\dhole{\dexp}{u}{\sigma}{\markname}}{\htau}
}

@@

\begin{figure}[!ht]
\judgbox{[\dexp_1 / x] \dexp_2 = \dexp_3}{Substitution}
\[
\begin{array}{lcl}
[\dexp_1 / x] c
&=&
c\\
{[\dexp_1 / x] x}
&=&
\dexp_1\\
{[\dexp_1 / x] y}
&=&
y\\
{[\dexp_1 / x] \dlam{y}{\htau}{\dexp_2}}
&=&
{\dlam{y}{\htau}{[\dexp_1 / x] \dexp_2}}
\\
{[\dexp_1 / x] \dap{e_2}{e_3}}
&=&
{\dapP{[\dexp_1 / x] e_2}{[\dexp_1 / x] e_3}}
\\
% {[\dexp_1 / x] \dinj{i}{e_2}}
% &=&
% {\dinj{i}{[\dexp_1 / x] e_2}}
% \\
% {[\dexp_1 / x] \dcase{e}{x}{e_x}{y}{e_y}}
% &=&
% {\dcase{[\dexp_1 / x]e}{x}{[\dexp_1 / x]e_x}{y}{[\dexp_1 / x]e_y}}
% \\
{[\dexp_1 / x]} \dehole{\mvar}{\subst}{\todo{??}}
&=&
\dehole{\mvar}{[\dexp_1 / x] \subst}{\todo{??}}
\\
{[\dexp_1 / x] \dhole{\dexp_2}{\mvar}{\subst}{\todo{??}}}
&=&
\dhole{ [ \dexp_1 / x ] \dexp_2 }{\mvar}{[\dexp_1 / x] \subst}{\todo{??}}
\end{array}
\]
\end{figure}

@@

\judgbox{\isIndet{\dexp}}{$\dexp$ is indeterminate}
\begin{mathpar}
\inferrule[IEHole]
{ }
{\isIndet{\dehole{\mvar}{\subst}{\todo{??}}}}

\inferrule[INEHole]
{\isFinal{\dexp}}
{\isIndet{\dhole{\dexp}{\mvar}{\subst}{\todo{??}}}}

@@

\judgbox{\isFinal{\dexp}}{$\dexp$ is final (it cannot be stepped)}
\begin{mathpar}
\Infer{FVal}
{\isValue{\dexp}}{\isFinal{\dexp}}
\and
\Infer{FIndet}
{\isIndet{\dexp}}{\isFinal{\dexp}}
%% No longer a rule: Errors are not indet anymore:
%% \and
%% \Infer{f-err}
%% {\isErr{\dexp}}{\isFinal{\dexp}}
\end{mathpar}

\todo{the judgmental forms are not consistent in these
rules; the contexts change. need to figure this out.}

\judgbox{\isErr{\Delta}{\dexp}}{Under $\Delta$, $\dexp$ contains a cast error}
\begin{mathpar}
\inferrule[ERNEHole~\todo{??}]
{\subst \sim_? \Gamma_2
\\
\isErr{\Delta; \Gamma_1, \Gamma_2}{\dexp}}
{\isErr{\Delta; \Gamma_1}{\dhole{\dexp}{\mvar}{\subst}}}

\inferrule[ERCastError~\todo{??}]
% XXX use macro type assignment judgement here
{\hasType{\Delta}{\todo{\emptyset}}{e}{\htau_2}\\
 \tinconsistent{\htau_1}{\htau_2}}
{\isErr{\Delta}{\dcast{\htau_1}{\dexp}}}

\inferrule[ERLam~\todo{??}]
% XXX use macro type assignment judgement here
{\isErr{\Delta; \Gamma, x:\htau_1}{\dexp}}
{\isErr{\Delta; \Gamma}{\halam{x}{\htau}{\dexp}}}

\\

\inferrule[ERAp1]
{
 \isErr{\Delta}{\dexp_1}
 %\\\
 %isFinal{\dexp_2}~\text{\todo{??}}
}
{\isErr{\Delta}{\dap{\dexp_1}{\dexp_2}}}

\inferrule[ERAp2]
{
 %\isIndet{\dexp_1}
 %\\
 \isErr{\Delta}{\dexp_2}
}
{\isErr{\Delta}{\dap{\dexp_1}{\dexp_2}}}

\inferrule[ERCast]
% XXX use macro type assignment judgement here
{\isErr{\Delta}{\dexp}}
{\isErr{\Delta}{\dcast{\htau_1}{\dexp}}}

\judgbox{\stepsToD{\Delta}{\dexp_1}{\dexp_2}}{under $\Delta$, $\dexp_1$
steps to $\dexp_2$}
\begin{mathpar}
\Infer{STHole}
{\stepsToD{\Delta}{\dexp_1}{\dexp_2} }
{\stepsToD{\Delta}{\dhole{\dexp_1}{\mvar}{\subst}{\todo{??}}}{\dhole{\dexp_2}{\mvar}{\subst}{\todo{??}}}}
\and
\Infer{STCast}
{\hasType{\Delta}{\emptyset}{\dexp}{\htau_2} \\
 \tconsistent{\tau_1}{\tau_2}}
{\stepsToD{\Delta}{\dcast{\htau_1}{\dexp}}{\dexp}}
\\
\Infer{STAp1}
{\stepsToD{\Delta}{\dexp_1}{\dexp_1'}}
{\stepsToD{\Delta}{\dap{\dexp_1}{\dexp_2}}{\dap{\dexp_1'}{\dexp_2}}}
\and
\Infer{STAp2}
{ \isFinal{\dexp_1} \\ \stepsToD{\Delta}{\dexp_2}{\dexp_2'}}
{\stepsToD{\Delta}{\dap{\dexp_1}{\dexp_2}}{\dap{\dexp_1}{\dexp_2'}}}
\and
\Infer{STAp$\beta$}
{ \isFinal{\dexp_2} }
{\stepsToD{\Delta}{\dapP{\dlam{x}{\htau}{\dexp_1}}{\dexp_2}}{ [\dexp_2/x]\dexp_1 }}
\end{mathpar}
\caption{Structural Dynamics}
\label{fig:isValue}
\label{fig:isIndet}
\label{fig:stepsTo}
\end{figure}

@@

\begin{theorem}[Progress]
If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either $\isValue{\dexp}$ or $\isIndet{\dexp}$ or $\isErr{\Delta}{\dexp}$ or $\stepsToD{\Delta}{\dexp}{\dexp'}$.
\end{theorem}

@@

\todo{complete progress + preservation}

@@

\todo{theorem that special casing the holes in EANEHole gives you ``the
most descriptive hole types'' for some sense of what that means -- they'd
all just be hole other wise}

>>>>

\inferrule[ESVar]{
  x : \htau \in \hGamma
}{
  \expandSyn{\hGamma}{x}{\htau}{x}{\emptyset}
}

@@

\inferrule[TAVar]{
  x : \htau \in \hGamma
}{
	\hasType{\Delta}{\hGamma}{x}{\htau}
}

@@

\inferrule[TAEHole]{
  \Dbinding{u}{\hGamma'}{\htau} \in \Delta\\
  \hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}
}{
  \hasType{\Delta}{\hGamma}{\dehole{u}{\sigma}{\markname}}{\htau}
}

\inferrule[TANEHole]{
  \Dbinding{u}{\hGamma'}{\htau} \in \Delta\\\\
  \hasType{\Delta}{\hGamma}{\dexp}{\htau'}\\
  \hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}
}{
  \hasType{\Delta}{\hGamma}{\dhole{\dexp}{u}{\sigma}{\markname}}{\htau}
}

@@

\begin{figure}[!ht]
\judgbox{[\dexp_1 / x] \dexp_2 = \dexp_3}{Substitution}
\[
\begin{array}{lcll}
[\dexp_1 / x] c
&=&
c\\
{[\dexp_1 / x] x}
&=&
\dexp_1\\
{[\dexp_1 / x] y}
&=&
y & (y \neq x)\\
{[\dexp_1 / x] \dlam{y}{\htau}{\dexp_2}}
&=&
{\dlam{y}{\htau}{[\dexp_1 / x] \dexp_2}}
& (y \neq x)
\\
{[\dexp_1 / x] \dap{d_2}{d_3}}
&=&
{\dapP{[\dexp_1 / x] d_2}{[\dexp_1 / x] d_3}}
\\
% {[\dexp_1 / x] \dinj{i}{e_2}}
% &=&
% {\dinj{i}{[\dexp_1 / x] e_2}}
% \\
% {[\dexp_1 / x] \dcase{e}{x}{e_x}{y}{e_y}}
% &=&
% {\dcase{[\dexp_1 / x]e}{x}{[\dexp_1 / x]e_x}{y}{[\dexp_1 / x]e_y}}
% \\
{[\dexp_1 / x]} \dehole{\mvar}{\subst}{m}
&=&
\dehole{\mvar}{[\dexp_1 / x] \subst}{m}
\\
{[\dexp_1 / x] \dhole{\dexp_2}{\mvar}{\subst}{m}}
&=&
\dhole{ [ \dexp_1 / x ] \dexp_2 }{\mvar}{[\dexp_1 / x] \subst}{m}
\\
{[\dexp_1 / x] \dcast{\htau}{\dexp}}
&=&
\dcast{\htau}{[\dexp_1 / x] \dexp}
\end{array}
\]
\end{figure}

@@

\judgbox{\isIndet{\dexp}}{$\dexp$ is indeterminate}
\begin{mathpar}
\inferrule[IEHole]
{ }
{\isIndet{\dehole{\mvar}{\subst}{\evaled}}}

\inferrule[INEHole]
{\isFinal{\dexp}}
{\isIndet{\dhole{\dexp}{\mvar}{\subst}{\evaled}}}

@@

\judgbox{\isFinal{\dexp}}{$\dexp$ is final (it cannot be stepped)}
\begin{mathpar}
\inferrule[FVal]
{\isValue{\dexp}}{\isFinal{\dexp}}
\and
\inferrule[FIndet]
{\isIndet{\dexp}}{\isFinal{\dexp}}
\end{mathpar}

\judgbox{\isErr{\Delta}{\dexp}}{$\dexp$ contains a cast error}
\begin{mathpar}
\inferrule[ECastError]{
  \hasType{\Delta}{\emptyset}{\dexp}{\htau_2}\\
  \tinconsistent{\htau_1}{\htau_2}
}{
  \isErr{\Delta}{\dcast{\htau_1}{\dexp}}
}

\inferrule[EConst]{
}{
  \isErr{\Delta}{\dcast{\tau}{\dconst}}
}

\inferrule[EAp1]{
  \isErr{\Delta}{\dexp_1}
}{
  \isErr{\Delta}{\dap{\dexp_1}{\dexp_2}}
}

\inferrule[EAp2]{
  \isErr{\Delta}{\dexp_2}
}{
  \isErr{\Delta}{\dap{\dexp_1}{\dexp_2}}
}

\inferrule[ENEHole]{
  \isErr{\Delta}{\dexp}
}{
  \isErr{\Delta}{\dhole{\dexp}{\mvar}{\subst}{m}}
}

\inferrule[ECastProp]{
  \isErr{\Delta}{\dexp}
}{
  \isErr{\Delta}{\dcast{\htau_1}{\dexp}}
}
\end{mathpar}
\caption{Auxiliary Judgments for the Dynamics}
\label{fig:isValue}
\label{fig:isIndet}
\label{fig:isFinal}
\label{fig:isErr}
\end{figure}

\begin{figure}
\judgbox{\stepsToD{\Delta}{\dexp_1}{\dexp_2}}{$\dexp_1$
steps to $\dexp_2$}
\begin{mathpar}
\inferrule[STEHoleEvaled]
{ }
{\stepsToD{\Delta}{\dehole{\mvar}{\subst}{\unevaled}}{\dehole{\mvar}{\subst}{\evaled}}}

\inferrule[STNEHoleStep]
{\stepsToD{\Delta}{\dexp_1}{\dexp_2} }
{\stepsToD{\Delta}{\dhole{\dexp_1}{\mvar}{\subst}{\unevaled}}{\dhole{\dexp_2}{\mvar}{\subst}{\unevaled}}}

\inferrule[STNEHoleEvaled]
{\isFinal{\dexp}}
{\stepsToD{\Delta}{\dhole{\dexp}{\mvar}{\subst}{\unevaled}}{\dhole{\dexp}{\mvar}{\subst}{\evaled}}}

\inferrule[STCast]
{
\isValue{\dexp}\\
\hasType{\Delta}{\emptyset}{\dexp}{\htau_2} \\
\tconsistent{\tau_1}{\tau_2}}
{\stepsToD{\Delta}{\dcast{\htau_1}{\dexp}}{\dexp}}

\inferrule[STApStep1]
{\stepsToD{\Delta}{\dexp_1}{\dexp_1'}}
{\stepsToD{\Delta}{\dap{\dexp_1}{\dexp_2}}{\dap{\dexp_1'}{\dexp_2}}}

\inferrule[STApStep2]
{ \isFinal{\dexp_1} \\ \stepsToD{\Delta}{\dexp_2}{\dexp_2'}}
{\stepsToD{\Delta}{\dap{\dexp_1}{\dexp_2}}{\dap{\dexp_1}{\dexp_2'}}}

\inferrule[STApBeta]
{ \isFinal{\dexp_2} }
{\stepsToD{\Delta}{\dapP{\dlam{x}{\htau}{\dexp_1}}{\dexp_2}}{ [\dexp_2/x]\dexp_1 }}
\end{mathpar}
\caption{Structural Dynamics}
\label{fig:stepsTo}
\end{figure}

\begin{figure}
$\arraycolsep=4pt\begin{array}{rllllll}
\mathsf{EvalCtx} & \evalctx & ::= &
  \evalhole ~\vert~
  \halam{x}{\htau}{\evalctx} ~\vert~
  \hap{\evalctx}{\dexp} ~\vert~
  \hap{\dexp}{\evalctx} ~\vert~
  \dhole{\evalctx}{\mvar}{\subst}{\markname} ~\vert~
  \dcast{\htau}{\evalctx}
\end{array}$

\judgbox{\isevalctx{\evalctx}}{$\evalctx$ is an evaluation context}
\begin{mathpar}
\inferrule[ECDot]{ }{
  \isevalctx{\evalhole}
}

\inferrule[ECLam]{
  \isevalctx{\evalctx}
}{
  \isevalctx{\halam{x}{\htau}{\evalctx}}
}

\inferrule[ECAp1]{
  \isFinal{\dexp}\\
  \isevalctx{\evalctx}
}{
  \isevalctx{\hap{\dexp}{\evalctx}}
}

\inferrule[ECAp2]{
  \isevalctx{\evalctx}
}{
  \isevalctx{\hap{\evalctx}{\dexp}}
}

\inferrule[ECNEHole]{
  \isevalctx{\evalctx}
}{
  \isevalctx{\dhole{\evalctx}{\mvar}{\subst}{\markname}}
}

\inferrule[ECCast]{
  \isevalctx{\evalctx}
}{
  \isevalctx{\dcast{\htau}{\evalctx}}
}
\end{mathpar}

\judgbox{\selectEvalCtx{d}{\evalctx}{d'}}{$d$ is the result of filling the
hole in $\evalctx$ with $d'$}
\begin{mathpar}
\inferrule[FDot]{ }{
  \selectEvalCtx{d}{\evalhole}{d}
}

%% \todo{d used to be c -- TODO IEV}

\inferrule[FLam]{
  \selectEvalCtx{d}{\evalctx}{d'}
}{
  \selectEvalCtx{\halam{x}{\htau}{d}}{\halam{x}{\htau}{\evalctx}}{d'}
}

\inferrule[FAp1]{
  \isFinal{d_1}\\
  \selectEvalCtx{d_2}{\evalctx}{d_2'}
}{
  \selectEvalCtx{\hap{d_1}{d_2}}{\hap{d_1}{\evalctx}}{d_2'}
}

\inferrule[FAp2]{
  \selectEvalCtx{d_1}{\evalctx}{d_1'}
}{
  \selectEvalCtx{\hap{d_1}{d_2}}{\hap{\evalctx}{d_2}}{d_1'}
}

%% I think these rules are not needed, if you make FDot work for any dhexp
%% d. The only reason they might be needed is because they're specified to
%% have check marks, not x marks, but i don't think that's important. i
%% think the more general FDot makes sense. -- TODO IEV
%%
%%
%% \inferrule[\todo{??}]{ }{
%%   \selectEvalCtx{\dehole{\mvar}{\subst}{\evaled}}{\evalhole}{\dehole{\mvar}{\subst}{\evaled}}
%% }

%% \inferrule[\todo{??}]{ }{
%%   \selectEvalCtx{\dhole{d}{\mvar}{\subst}{\evaled}}{\evalhole}{\dhole{d}{\mvar}{\subst}{\evaled}}
%% }

\inferrule[\todo{??}]{
  \selectEvalCtx{d}{\evalctx}{d'}
}{
  \selectEvalCtx{\dhole{d}{\mvar}{\subst}{\unevaled}}{\dhole{\evalctx}{\mvar}{\subst}{\unevaled}}{d'}
}

\inferrule[\todo{??}]{
  \isFinal{d}
}{
  \selectEvalCtx{\dhole{d}{\mvar}{\subst}{\unevaled}}{\evalhole}{\dhole{d}{\mvar}{\subst}{\unevaled}}
}

\inferrule[FCast]{
  \selectEvalCtx{d}{\evalctx}{d'}
}{
  \selectEvalCtx{\dcast{\htau}{d}}{\dcast{\htau}{\evalctx}}{d'}
}
\end{mathpar}

\judgbox{\reducesE{\Delta}{\dexp_1}{\dexp_2}}{$\dexp_1$
transitions to $\dexp_2$}
\begin{mathpar}
\inferrule[ITLam]{
  \isFinal{d_2}
}{
  \reducesE{\Delta}{\hap{(\halam{x}{\htau}{d})}{d_2}}{[d_2/x]d}
}

\inferrule[ITCast]{
  \isFinal{d}\\
  \hasType{\Delta}{\emptyset}{d}{\tau_2}\\
  \tconsistent{\tau_1}{\tau_2}
}{
  \reducesE{\Delta}{\dcast{\htau_1}{d}}{d}
}

\inferrule[ITEHole]{ }{
  \reducesE{\Delta}{\dehole{\mvar}{\subst}{\unevaled}}{\dehole{\mvar}{\subst}{\evaled}}
}

\inferrule[ITNEHole]{
  \isFinal{d}
}{
  \reducesE{\Delta}{\dhole{d}{\mvar}{\subst}{\unevaled}}{\dhole{d}{\mvar}{\subst}{\evaled}}
}
\end{mathpar}

\judgbox{\stepsToD{\Delta}{\dexp_1}{\dexp_2}}{$\dexp_1$ steps to $\dexp_2$}
\begin{mathpar}
\inferrule[Step]{
  \selectEvalCtx{d}{\evalctx}{d_0}\\
  \reducesE{\Delta}{d_0}{d_0'}\\
  \selectEvalCtx{d'}{\evalctx}{d_0'}
}{
  \stepsToD{\Delta}{\dexp}{\dexp'}
}
\end{mathpar}
\caption{Contextual Dynamics}
\label{fig:contextual-dynamics}
\end{figure}

@@

\begin{theorem}[Progress]
If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either
$\isValue{\dexp}$ or $\isIndet{\dexp}$ or $\isErr{\Delta}{\dexp}$ or
$\stepsToD{\Delta}{\dexp}{\dexp'}$ for some $\dexp'$.
\end{theorem}

@@

\begin{theorem}[Complete Progress \todo{??}]
If $\hcomplete{\hexp}$ and
$\expandSyn{\emptyset}{\hexp}{\htau}{\dexp}{\Delta}$ then either
$\isValue{\dexp}$ or $\stepsToD{\Delta}{\dexp}{\dexp'}$ for some $\dexp'$.

\end{theorem}

\begin{theorem}[Complete Preservation \todo{??}]
If $\hcomplete{\hexp}$ and
$\expandSyn{\emptyset}{\hexp}{\htau}{\dexp}{\Delta}$ and
$\stepsToD{\Delta}{\dexp}{\dexp'}$ then
$\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$
\end{theorem}

\begin{theorem}[Commutativity \todo{state this}]
\end{theorem}

\begin{theorem}[Maximum Informativity]
If the expansion produces $t1$, and there exists another possible type choice
$t2$, then $t1 \sim t2$ and $t1 JOIN t2 = t1$
\end{theorem}\footnote{idea is that special casing the holes in EANEHole gives you ``the
most descriptive hole types'' for some sense of what that means -- they'd
all just be hole other wise. from Matt:
\begin{quote}
It sounds like we need a something akin to an abstract domain (a lattice),
where hole has the least information, and a fully-defined type (without
holes) has the most information.  You can imagine that this lattice really
expands the existing definition we have of type consistency, which is
merely the predicate that says whether two types are comparable
(“join-able”) in this lattice.  lattice join is the operation that goes
through the structure of two (consistent) types, and chooses the structure
that is more defined (i.e., non-hole, if given the choice between hole and
non-hole).

The rule choosenonhole below is the expansion of this consistency rule that
we already have (hole consistent with everything)
\end{quote}}

\begin{verbatim}
t not hole
-------------------- :: choose-non-hole
hole JOIN t  = t
\end{verbatim}
\begin{verbatim}
------------ :: hole-consistent-with-everything
hole ~ t
\end{verbatim}

@@
